### 2014-10-02 Studies/2011_San_Francisco/Report/San_Francisco_Fire.tex
San Francisco: Add figures and discussion to flow path and hazard sections

The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the stairwell. The temperature of the gases in the interior stairwell was estimated to be in excess of 600~$^{\circ}$C (1100~$^{\circ}$F), which exceeds the Class~III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F). Unfortunately, two E26 firefighters were located in the flow path between the basement and the doors on the front side of the structure. After a Mayday call, both firefighters were removed from the structure and immediate medical treatment was provided. The two firefighters were transported to the local medical center where the lieutenant was pronounced dead and the fire fighter/paramedic died two days later.

@@

In this figure, the pressure profile within the high hazard areas is shown via two snapshots in time to illustrate the change in the interior conditions. The first snapshot is shown at a time of 340~s into the simulation, which is 2~s before the first rear basement window failed, and the second snapshot is shown at a time of 402~s into the simulation, which is 60~s after the first rear basement window failed. The pressure rise in the basement ranges from a 5~Pa ($7.3 \times 10^{-4}$~psi) over-pressure at mid-level height to greater than 10~Pa ($1.5 \times 10^{-3}$~psi) at the ceiling.

@@

The velocity vectors shown in Fig.~\ref{fig:smv_velocity_vectors} indicate the direction and magnitude of flow velocities within the structure before and after the rear basement window failures. Based on Figs.~\ref{fig:smv_velocity} and \ref{fig:smv_velocity_vectors}, the maximum velocity in the interior stairwell was approximately 9~m/s (20 mph).

@@

The temperature vectors shown in Fig.~\ref{fig:smv_temperature_vectors} indicate the gas temperatures and direction of flow within the structure before and after the rear basement window failures. Based on the results shown in Figs.~\ref{fig:smv_temperature} and \ref{fig:smv_temperature_vectors} after the rear basement windows began to fail, the maximum temperature in the basement was in excess of 800~$^{\circ}$C (1500~$^{\circ}$F), and the maximum temperature in the interior stairwell was in excess of 600~$^{\circ}$C (1100~$^{\circ}$F).

@@

The NIOSH report~\cite{NIOSH:Bowyer2} indicated that the interior stairwell on the front side of the structure was used to determine that the fire was located in the basement. From the simulation results (Section~\ref{sec:temperature}), the conditions in the interior stairwell were initially tenable. After the rear basement windows failed, the simulation results indicate a high-hazard area (temperatures greater than 260~$^{\circ}$C or 500~$^{\circ}$F) in the stairwell near the laundry room landing area.

@@

\begin{figure}[!ht]
\includegraphics[width=5.0in]{../Figures/Flow_Path_Top}
\caption{Summary of simulated flow path in the stairwell on the basement floor and first floor after the rear basement window failures.}
\label{fig:flow_path_top}
\end{figure}

@@

A person is susceptible to second-degree burn injuries if exposed to temperatures greater than 55~$^{\circ}$C (130~$^{\circ}$F)~\cite{contactburn}. Although firefighters wear protective gear, gear only offers a finite amount of protection. The polycarbonate in a self contained breathing apparatus begins to soften when the material temperature reaches approximately 140~$^{\circ}$C (284~$^{\circ}$F)~\cite{mensch2011emergency}. Structural fire fighting coats and pants are tested to withstand temperatures of 260~$^{\circ}$C (500~$^{\circ}$F)~\cite{nfpa2013standard}. Prolonged exposure to elevated temperatures can result in a significant amount of heat transferred to the firefighter, putting him/her at risk. Exposure of equipment to temperatures of 260~$^{\circ}$C (500~$^{\circ}$F) represents a Class~III exposure~\cite{Donnelly2006}. Firefighters are at increased risk levels when encountering Class~III exposure conditions for more than 5~minutes~\cite{Donnelly2006}. Based on the simulation results shown in Section~\ref{sec:temperature}, the simulated flow path temperatures in the interior stairwell were in excess of 600~$^{\circ}$C (1100~$^{\circ}$F) and the flow velocities were approximately 9~m/s (20 mph).

% The temperature of the gases that flowed into the interior stairwell was estimated to be approximately XXX~$^{\circ}$C (XXX~$^{\circ}$F), which exceeds the Class~III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F).

% Prior to the interior door failure, the elevated temperatures on the second floor were limited to the rear porch, as shown in Fig.~\ref{fig:temp_top_159}. The temperatures were greatest at the back-left corner where there is sufficient oxygen due to the section of porch wall that had burned away and allowed additional oxygen to mix with the fuel-rich fire gases. Five seconds after the interior door failed, hot gases traveled the length of the interior hallway and approached the doorway into the kitchen, which was approximately 4.5~m (14.8~ft) away from the source fire, as shown in Fig.~\ref{fig:temp_top_165}. Within 60~s of the interior door failure, the hallway temperatures were approximately at or above 260~$^{\circ}$C (500~$^{\circ}$F) throughout the hallway, as shown in Fig.~\ref{fig:temp_top_220}. These results indicate that the high temperature gases seen 5~s after failure were not just a burst but rather the front of a sustained exhaust gas flow from the fire burning in the enclosed rear porch. The conditions in this hallway changed from tenable to high-hazard very rapidly following the door failure. Figure~\ref{fig:temp_top_220} also shows that the temperatures in the kitchen were at or below 90~$^{\circ}$C (194~$^{\circ}$F), and the temperatures at the top of the stairs were at or below 70~$^{\circ}$C (158~$^{\circ}$F). At these locations within the structure, the exhaust gases in the flow path were cooled and diluted by the inflowing ambient gas. These simulated temperatures are consistent with the post-incident conditions that were documented in the kitchen and second floor stairwell.

% Insert top-down figures from SMV showing laundry room landing area temperatures.

% \begin{figure}[!ht]
% \includegraphics[width=6.5in]{../Figures/west_50th_baseline_top_159_6ft}
% \input{../Scripts/colorbar_std} 
% \input{../Scripts/colorbar_metric}
% \caption[Temperature contours on the second floor, 1~s prior to interior door failure.]
% {Temperature contours on the second floor, 1~s prior to interior door failure, 1.83~m (6~ft) above the floor.}
% \label{fig:temp_top_159}
% \end{figure}

% \begin{figure}[!ht]
% \includegraphics[width=6.5in]{../Figures/west_50th_baseline_top_165_6ft}
% \input{../Scripts/colorbar_std} 
% \input{../Scripts/colorbar_metric}
% \caption[Temperature contours on the second floor, 5~s after interior door failure.]
% {Temperature contours on the second floor, 5~s after interior door failure, 1.83~m (6~ft) above the floor.}
% \label{fig:temp_top_165}
% \end{figure}

% \begin{figure}[!ht]
% \includegraphics[width=6.5in]{../Figures/west_50th_baseline_top_220_6ft}
% \input{../Scripts/colorbar_std} 
% \input{../Scripts/colorbar_metric}
% \caption[Temperature contours on the second floor, 60~s after interior door failure.]
% {Temperature contours on the second floor, 60~s after interior door failure, 1.83~m (6~ft) above the floor.}
% \label{fig:temp_top_220}
% \end{figure}

@@

The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows failed, a flow path was established between the basement living room area and the doors located on the front side of the structure (the front door and the garage door). The rear basement window failures resulted in a rapid change in the conditions within the flow path. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the stairwell. The temperature of the gases in the interior stairwell was estimated to be in excess of 600~$^{\circ}$C (1100~$^{\circ}$F), which exceeds the Class~III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F).

>>>>

The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the stairwell. The temperature of the gases in the interior stairwell was estimated to be in excess of 760~$^{\circ}$C (1400~$^{\circ}$F), which exceeds the Class~III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F). Unfortunately, two E26 firefighters were located in the flow path between the basement and the doors on the front side of the structure. After a Mayday call, both firefighters were removed from the structure and immediate medical treatment was provided. The two firefighters were transported to the local medical center where the lieutenant was pronounced dead and the fire fighter/paramedic died two days later.

@@

In this figure, the pressure profile within the high hazard areas is shown via two snapshots in time to illustrate the change in the interior conditions. The first snapshot is shown at a time of 340~s into the simulation, which is 2~s before the first rear basement window failed, and the second snapshot is shown at a time of 402~s into the simulation, which is 60~s after the first rear basement window failed. After the rear basement windows failed, the pressure rise in the basement ranged from a 5~Pa ($7.3 \times 10^{-4}$~psi) over-pressure at mid-level height to greater than 10~Pa ($1.5 \times 10^{-3}$~psi) at the ceiling.

@@

The velocity vectors shown in Fig.~\ref{fig:smv_velocity_vectors} indicate the direction and magnitude of flow velocities within the structure before and after the rear basement window failures. Based on Figs.~\ref{fig:smv_velocity} and \ref{fig:smv_velocity_vectors}, the velocity in the interior stairwell was approximately 4~m/s (9~mph) before the rear basement windows failed and approximately 9~m/s (20~mph) after the rear basement windows failed.

@@

The temperature vectors shown in Fig.~\ref{fig:smv_temperature_vectors} indicate the gas temperatures and direction of flow within the structure before and after the rear basement window failures. Based on the results shown in Figs.~\ref{fig:smv_temperature} and \ref{fig:smv_temperature_vectors}, 60~s after the rear basement windows began to fail, the temperature in the basement was in excess of 800~$^{\circ}$C (1500~$^{\circ}$F), and the temperature in the interior stairwell was in excess of 540~$^{\circ}$C (1000~$^{\circ}$F).

@@

The NIOSH report~\cite{NIOSH:Bowyer2} indicated that the interior stairwell on the front side of the structure was used to determine that the fire was located in the basement. From the simulation results (Section~\ref{sec:temperature}), the conditions in the interior stairwell were initially tenable. After the rear basement windows failed, the simulation results indicate a high-hazard area that exceeded the conditions of a Class~III exposure (temperatures greater than 260~$^{\circ}$C or 500~$^{\circ}$F) in the stairwell near the laundry room landing area.

@@

\begin{figure}[!ht]
\includegraphics[width=6.5in]{../Figures/Flow_Path_Top_Vel_462_s_First_Floor}
\includegraphics[width=6.5in]{../Figures/Flow_Path_Top_Vel_462_s_Basement_Floor}
\caption{Summary of simulated flow path in the stairwell on the first floor and basement floor 120~s after the first rear basement window failure. The arrows indicate the direction of flow. Two arrows located at a door or window indicate bi-directional flow at a vent opening. The maximum velocity (red) corresponds to 9~m/s (20~mph).}
\label{fig:flow_path_top}
\end{figure}

@@

A person is susceptible to second-degree burn injuries if exposed to temperatures greater than 55~$^{\circ}$C (130~$^{\circ}$F)~\cite{contactburn}. Although firefighters wear protective gear, gear only offers a finite amount of protection. The polycarbonate in a self contained breathing apparatus begins to soften when the material temperature reaches approximately 140~$^{\circ}$C (284~$^{\circ}$F)~\cite{mensch2011emergency}. Structural fire fighting coats and pants are tested to withstand temperatures of 260~$^{\circ}$C (500~$^{\circ}$F)~\cite{nfpa2013standard}. Prolonged exposure to elevated temperatures can result in a significant amount of heat transferred to the firefighter, putting him/her at risk. Exposure of equipment to temperatures of 260~$^{\circ}$C (500~$^{\circ}$F) represents a Class~III exposure~\cite{Donnelly2006}. Firefighters are at increased risk levels when encountering Class~III exposure conditions for more than 5~minutes~\cite{Donnelly2006}.

Figure~\ref{fig:smv_flow_path_top_1} shows a top view of the temperature contours (at a height of 1.2~m from the floor) on the first floor at a time of 2~s before the first rear basement window failed. At this time, the temperature of the gases in the interior stairwell exceeded 200~$^{\circ}$C (400~$^{\circ}$F).

\begin{figure}[!ht]
\includegraphics[width=4.3in]{../Figures/SMV_Top_Temp_340_s}
\input{../Figures/colorbar_metric_temp}
\input{../Figures/colorbar_std_temp}
\caption{Top view of simulated temperature contours on the first floor, 1.2~m from the floor (mid-level height), 2~s before the first rear basement window failure (342~s).}
\label{fig:smv_flow_path_top_1}
\end{figure}

Figure~\ref{fig:smv_flow_path_top_2} shows a top view of the temperature contours (at a height of 1.2~m from the floor) on the first floor at a time of 60~s and 120~s after the first rear basement window failed. At a time of 60~s after the first rear basement window failed, the temperature of the gases in the interior stairwell exceeded 540~$^{\circ}$C (1000~$^{\circ}$F), which exceeds the Class~III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F). As more rear basement windows failed and provided ambient air to the growing basement fire, the temperatures in the interior stairwell continued to rise. At a time of 120~s after the first rear basement window failed, the temperature of the gases in the interior stairwell exceeded 760~$^{\circ}$C (1400~$^{\circ}$F), which exceeds the Class~III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F).

\begin{figure}[!ht]
\includegraphics[width=4.3in]{../Figures/SMV_Top_Temp_402_s}
\input{../Figures/colorbar_metric_temp}
\input{../Figures/colorbar_std_temp}
\includegraphics[width=4.3in]{../Figures/SMV_Top_Temp_462_s}
\input{../Figures/colorbar_metric_temp}
\input{../Figures/colorbar_std_temp}
\caption{Top view of simulated temperature contours on the first floor, 1.2~m from the floor (mid-level height), 60~s (top) and 120~s (bottom) after the first rear basement window failure (342~s).}
\label{fig:smv_flow_path_top_2}
\end{figure}

Based on these results and the simulation results shown in Sections~\ref{sec:velocity} and \ref{sec:temperature}, after the rear basement windows failed, the simulated flow path temperatures in the interior stairwell were in excess of 760~$^{\circ}$C (1400~$^{\circ}$F) and the flow velocities were approximately 9~m/s (20 mph). The conditions in the interior stairwell changed from tenable to high-hazard very rapidly following the rear basement window failure. Within the structure, the exhaust gases in the flow path were moving towards the doors located on the front side of the structure via the interior stairwell. These simulated temperatures are consistent with the post-incident conditions that were documented in the interior stairwell.

@@

The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows failed, a flow path was established between the basement living room area and the doors located on the front side of the structure (the front door and the garage door). The rear basement window failures resulted in a rapid change in the conditions within the flow path. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the stairwell. The temperature of the gases in the interior stairwell was estimated to be in excess of 760~$^{\circ}$C (1400~$^{\circ}$F), which exceeds the Class~III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F).
